Geometrization of the local Langlands correspondence for algebraic tori

### 2014-05-21 geometrization.tex
Title

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[Geometrization of quasicharacters]{Geometrization of quasicharacters of algebraic tori over local fields}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
%\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
%\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\qcs}[1]{{\mathcal{#1}}}
\newcommand{\gqcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gqcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\qcs{#1}^\dagger}
%% Categories
\newcommand{\QC}{{\mathcal{Q\hskip-0.8pt C}}}
\newcommand{\QCb}{{\QC_0}}
\newcommand{\QCf}{{\QC_f}}
\newcommand{\QCiso}[1]{\QC(#1)_{/\text{iso}}}
\newcommand{\QCbiso}[1]{\QCb(#1)_{/\text{iso}}}
\newcommand{\QCfiso}[1]{\QCf(#1)_{/\text{iso}}}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{T_{#1}}
%% amsmath replacements for \atop
\newcommand{\latop}[2]{\genfrac{}{}{0pt}{0}{#1}{#2}}
\newcommand{\partop}[2]{\genfrac{(}{)}{0pt}{0}{#1}{#2}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{\bar{g}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gqcs{L} \tight{0}{\boxtimes}{0} \gqcs{L}$}}
\newcommand{\ExE}{\text{$\qcs{E}\tight{0}{\boxtimes}{0}\qcs{E}$}}
\newcommand{\bExE}{\text{$\gqcs{E}\tight{0}{\boxtimes}{0}\gqcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Margin notes
\newcommand\Clifton[1]{\marginpar{\smaller\smaller CC: #1}}
\newcommand\David[1]{\marginpar{\smaller\smaller DR: #1}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes locally of finite type over finite fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 As an application, we find a categorical transfer of certain characters between characteristic-$p$ tori and characteristic-$0$ tori.
  \end{abstract}

@@

We explain how to geometrize quasicharacters of algebraic tori over local fields.  
%
Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$. Then $\GN{T}_n$ is a commutative group scheme locally of finite type over $\Fq$ with finitely generated geometric component group, and as such, it carries the category $\QCiso{\GN{T}_n}$ of quasicharacter sheaves, introduced in \cite{cunningham-roe:dictionary}. The main result in this paper, Theorem~\ref{thm:application}, provides a functorial translation of characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$: 
\[
\QCiso{\GN{T}_n} 
\iso
\Hom_{< n}(T(K),\EEx).
\]  
We also describe a pro-group scheme $\GN{T}$ 
simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

@@

Let $\Fq$ be a finite field and let $K$ be a non-archimedean local field with residue field $\Fq$.
We place no restriction on the characteristic of $K$. 
We write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$. 
Let $T$ be an algebraic torus over $K$; the filtrations defined by Moy-Prasad and Yu \cite{yu:03a}*{\S 5}
allow us to define the depth of character $\chi$ as the infimum of the $r$ so that $\chi$ vanishes on $T(K)_r$.
We write $\Hom_{< n}(T(K),\EEx)$ for the group of characters of $T(K)$ depth less than $n$ 
and $\Hom_{\text{qc}}(T(K),\EEx)$ for the group of characters of $T(K)$ of arbitrary finite depth;
we call these quasicharacters of $T(K)$, following \cite{tate:67a}*{2.3}.

@@

The multiplication maps $m_n : \GN{T}_n \times \GN{T}_n \to \GN{T}_n$ induce maps of affine schemes
$m_n^{i,j} : \GN{T}_n^i \times \GN{T}_n^j \to \GN{T}_n^{i+j},$ inducing maps $m^{i,j} : \GN{T}^i \times \GN{T}_

\end{proof}

\section{The function--sheaf dictionary}\label{sec:qcs}

@@

Let $G$ be a smooth commutative group schemes locally of finite type over $\Fq$ for which the geometric component group is finitely generated. A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; see \cite{cunningham-roe:dictionary}.
%
Quasicharacter sheaves on $G$ form a rigid monoidal category, denoted by $\QC(G)$
and the trace of Frobenius
\begin{equation}\label{snake}
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
\end{equation}
  is an isomorphism of groups.

\section{Application to quasicharacters of bounded depth} \label{sec:bdchar}

We now apply the results of Section~\ref{sec:qcs} to characters of $T(K)$.

@@

\section{A function--sheaf dictionary for tori over local fields} \label{sec:quasichar}

@@

\end{document}

>>>>

%%%%%%%%%%%%%%%%%% HEADING %%%%%%%%%%%%%%%%%%
\title[Geometrization of the local Langlands correspondence for tori]{Geometrization of the local Langlands correspondence for algebraic tori}
\date{\today}
\author{Clifton Cunningham}
\email{cunning@math.ucalgary.ca}
\address{University of Calgary}
\author{David Roe}
\email{roed.math@gmail.com}
\address{Pacific Institute for the Mathematical Sciences at the University of Calgary}
%\classification{14L15 (primary), 11G25, 14F05, 14D24, 11S31 (secondary)}
%\keywords{quasicharacters, quasicharacter sheaves, algebraic tori, Greenberg transforms, N\'eron models, local fields, group schemes}

@@

%%%%%%%%%%%% MISCELLANEOUS OPERATORS %%%%%%%%%%%%
\newcommand{\sheafHom}{{\mathscr{H}\hskip-4pt{\it o}\hskip-2pt{\it m}}}
\newcommand{\abs}[1]{{\vert #1 \vert}}
\newcommand{\ceq}{{\, :=\, }}
\newcommand{\tq}{{\ \vert\ }}
\newcommand{\iso}{{\ \cong\ }}
\newcommand{\dual}[1]{{\hat{#1}}}
%% Limits
\newcommand{\invlim}[1]{\lim\limits_{\overleftarrow{#1}}}
\newcommand{\dirlim}[1]{\lim\limits_{\overrightarrow{#1}}}
\newcommand{\limit}[1]{\mathop{\textsc{lim}}\limits_{#1}}
\newcommand{\colimit}[1]{\mathop{\textsc{colim}}\limits_{#1}}
%% Fonts for quasicharacter sheaves
\newcommand{\qcs}[1]{{\mathcal{#1}}}
\newcommand{\gqcs}[1]{{\mathcal{\bar #1}}}
\newcommand{\dualgcs}[1]{\gqcs{#1}^\dagger}
\newcommand{\dualcs}[1]{\qcs{#1}^\dagger}
%% Categories
\newcommand{\QC}{{\mathcal{Q\hskip-0.8pt C}}}
\newcommand{\QCb}{{\QC_0}}
\newcommand{\QCf}{{\QC_f}}
\newcommand{\QCiso}[1]{\QC(#1)_{/\text{iso}}}
\newcommand{\QCbiso}[1]{\QCb(#1)_{/\text{iso}}}
\newcommand{\QCfiso}[1]{\QCf(#1)_{/\text{iso}}}
\newcommand{\trFrob}[1]{t_{#1}}
\newcommand{\TrFrob}[1]{T_{#1}}
%% amsmath replacements for \atop
\newcommand{\latop}[2]{\genfrac{}{}{0pt}{0}{#1}{#2}}
\newcommand{\partop}[2]{\genfrac{(}{)}{0pt}{0}{#1}{#2}}
%% Labeled items
\makeatletter
\newcommand{\labitem}[2]{
\def\@itemlabel{\textbf{#1}}
\item
\def\@currentlabel{#1}\label{#2}}
\makeatother
%% Shorthand for bars
\renewcommand{\bf}{\bar{f}}
\newcommand{\bg}{\bar{g}}
\newcommand{\bm}{\bar{m}}
\newcommand{\bG}{\bar{G}}
\newcommand{\bH}{\bar{H}}
%% Spacing control
\newcommand{\tight}[3]{\hspace{-#1pt}{#2}\hspace{-#3pt}}
\newcommand{\GxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\bGxG}{\text{$\bar{G} \tight{1}{\times}{1} \bar{G}$}}
\newcommand{\bfxf}{\text{$\bar{f} \tight{1}{\times}{1} \bar{f}$}}
\newcommand{\GxxG}{\text{$G \tight{1}{\times}{1} G$}}
\newcommand{\LxL}{\text{$\gqcs{L} \tight{0}{\boxtimes}{0} \gqcs{L}$}}
\newcommand{\ExE}{\text{$\qcs{E}\tight{0}{\boxtimes}{0}\qcs{E}$}}
\newcommand{\bExE}{\text{$\gqcs{E}\tight{0}{\boxtimes}{0}\gqcs{E}$}}
\newcommand{\AxA}{\text{$A \tight{1}{\times}{1} A$}}
\newcommand{\BxB}{\text{$B \tight{1}{\times}{1} B$}}
\newcommand{\GzxGz}{\text{$G^0 \tight{1}{\times}{1} G^0$}}
%% Margin notes
\newcommand\Clifton[1]{\marginpar{\smaller\smaller CC: #1}}
\newcommand\David[1]{\marginpar{\smaller\smaller DR: #1}}
%% Hyphenation override
\hyphenation{quasi-character}

@@

\begin{abstract}
For any non-archimedean local field $K$
  and any algebraic torus $T$ over $K$, we use the
  function--sheaf dictionary for smooth group schemes over finite fields
  to geometrize quasicharacters of $T(K)$
  as quasicharacter sheaves on the Greenberg transform of the Néron
  model of $T$.
 We offer two applications of this geometrization:
 a categorical transfer of quasicharacters between characteristic-$p$ tori and
 characteristic-$0$ tori; and a geoemtric proof of the local Langlands correspondence for algebraic tori.
  \end{abstract}

@@

In this paper we explain how to geometrize quasicharacters of algebraic tori over local fields.
%
Let $K$ be a non-archimedian local field
with finite residue field $\Fq$; we place no restriction on the characteristic of $K$.
Write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be a torus defined over $K$.  The N\'eron model $\TT$ of $T$
is a smooth group scheme over $\OK$ with generic fibre $T$ and a canonical isomorphism
$\TT(\OK) \cong T(K)$.  For each positive integer $n$, let $\GN{T}_n$ be the
Greenberg transform of $\TT \times_{\Spec{\OK}} \Spec{\OK/\pK^n}$. Then $\GN{T}_n$ is a smooth commutative group scheme over $\Fq$; as such, it carries the category $\QC(\GN{T}_n)$ of quasicharacter sheaves, introduced in \cite{cunningham-roe:dictionary}. The main result in this paper, Theorem~\ref{thm:application}, provides a functorial translation of characters $T(K)$ of depth less than $n$ into quasicharacter sheaves on $\GN{T}_n$:
\[
\QCiso{\GN{T}_n}
\iso
\Hom_{< n}(T(K),\EEx).
\]
We also describe a pro-group scheme $\GN{T}$
simultaneously geometrizing all quasicharacters ({\it i.e.}, finite-depth characters) of $T(K)$.

\[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr} & \\
\Hom_{\text{qc}}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell)
\end{tikzcd}
\]

@@

Let $\Fq$ be a finite field and let $K$ be a non-archimedean local field with residue field $\Fq$.
We place no restriction on the characteristic of $K$.
We write $\OK$ for the ring of integers of $K$ and $\pK$ for the maximal ideal in $\OK$.
Let $T$ be an algebraic torus over $K$; the filtrations defined by Moy-Prasad and Yu \cite{yu:03a}*{\S 5}
allow us to define the depth of character $\chi$ as the infimum of the $r$ so that $\chi$ vanishes on $T(K)_r$.
We write
\[
\Hom_{< n}(T(K),\EEx)
\]
for the group of characters of $T(K)$ depth less than $n$
and
\[
\Hom_{\text{qc}}(T(K),\EEx)
\]
for the group of characters of $T(K)$ of arbitrary finite depth;
we call these quasicharacters of $T(K)$, following \cite{tate:67a}*{2.3}.

@@

The multiplication maps $m_n : \GN{T}_n \times \GN{T}_n \to \GN{T}_n$ induce maps of affine schemes
$m_n^{i,j} : \GN{T}_n^i \times \GN{T}_n^j \to \GN{T}_n^{i+j},$ inducing maps $m^{i,j} : \GN{T}^i \times \GN{T}$ [under construction]
\end{proof}

\section{The function--sheaf dictionary for quasicharacters of bounded depth}\label{sec:qcs}

@@

Let $G$ be a smooth commutative group schemes locally of finite type over $\Fq$ for which the geometric component group is finitely generated. A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; see \cite{cunningham-roe:dictionary}.
%
Quasicharacter sheaves on $G$ form a rigid monoidal category, denoted by $\QC(G)$
and the trace of Frobenius
\begin{equation}\label{snake}
  \trFrob{G} : \QCiso{G} \to G(\Fq)^*
\end{equation}
  is a functorial isomorphism of groups \cite{cunningham-roe:dictionary}*{Thm. 11.1}.

%\section{Application to quasicharacters of bounded depth} \label{sec:bdchar}

%We now apply the results of Section~\ref{sec:qcs} to characters of $T(K)$.

@@

\section{A function--sheaf dictionary for quasicharacters} \label{sec:quasichar}

@@

\end{document}
%sagemathcloud={"zoom_width":90}
